---
title: "GeneDupeRExplorations"
output: html_document
date: '2022-03-17'
---

```{r}
library(ggplot2)
setwd("/Users/cameronhaynes/Documents/VSCodeGitProjects/AvidaGeneDupe/experiments/2022-3-29-PaperDuplication/TidiedData")
data <- read.csv("FinalDominantTaskCounts.csv")
names(data)
ggplot(data,aes(x=Treatment,y=Task.Count,color=Treatment,fill=Treatment)) + #Add color=Replicate with a replicate column to plot each curve seperately 
  geom_violin() +
  geom_jitter(alpha = 0.2) +
  theme_minimal() +
  theme(legend.position = 'none')#+ facet_grid(Treatment~.) # Uncomment to plot faceted by treatment

ggsave("ModifiedReplicationFinalDominantTaskCountViolin.pdf")
```
```
```{r}
library(ggplot2)
setwd("/Users/cameronhaynes/Documents/VSCodeGitProjects/AvidaGeneDupe/experiments/2022-2-4-PaperDuplicationWMod/TidiedData")
data <- read.csv("TaskCountTimeCourse.csv")
names(data)
ggplot(data, aes(x=Time,y=Tasks.Present,color=Treatment,fill=Treatment)) +
  stat_summary(geom="line",fun=mean)

```
```{r}
library(ggplot2)
setwd("/Users/cameronhaynes/Documents/VSCodeGitProjects/AvidaGeneDupe/experiments/2022-3-29-PaperDuplication/TidiedData")
data <- read.csv("TaskCountTimeCourse.csv")
names(data)
ggplot(data, aes(x=Time,y=Tasks.Present,color=Treatment,fill=Treatment)) +
  stat_summary(geom="line",fun=mean)
```

